---
title: "Plotting with the Linköping University ggplot2 theme"
author: "Nils Fahrni"
date: "09-25-2025"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Plotting with the Linköping University ggplot2 theme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4,
  out.width = "70%"
)
library(lab4)
library(ggplot2)
```

# Overview

This vignette introduces `liu_theme()`, a minimalistic **ggplot2** theme inspired by Linköping University's visual style. It builds on `theme_minimal()` and adjusts background, grid, and titles for clean, legible charts out of the box.

```{r}
liu_theme <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      plot.background = ggplot2::element_rect(fill = "#d1f4f6"),
      plot.title = ggplot2::element_text(face = "bold"),
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major = ggplot2::element_line(colour = "#54d8e0")
    )
}
```

> If you are using this inside a package, you don't need to redefine it here—load your package instead, e.g. `library(yourpackagename)`.

# Quick demo

```{r}
library(ggplot2)

p <- ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point() +
  labs(
    title = "Fuel economy by engine size",
    subtitle = "EPA data (ggplot2::mpg)",
    x = "Engine displacement (litres)",
    y = "Highway MPG",
    colour = "Class"
  )

# Default ggplot2 look
p
```

Apply `liu_theme()`:

```{r}
p + liu_theme()
```

# Set as default (optional)

To use the theme for all subsequent plots in your session:

```{r}
op_theme <- theme_set(liu_theme())
# Your plots now use liu_theme() by default
ggplot(mtcars, aes(wt, mpg)) + geom_point() + labs(title = "Default now: liu_theme()")
```

# Customization tips

Because `liu_theme()` returns a standard ggplot2 theme, you can layer modifications with `+ theme(...)`:

```{r}
p + liu_theme() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(size = 14),
    panel.grid.major = element_line(linetype = 3)
  )
```

You can also combine it with scale helpers without conflict:

```{r}
p + liu_theme() +
  scale_colour_brewer(palette = "Dark2")
```

# Color choices and accessibility

- **Plot background** `#d1f4f6`: light, low-ink background that frames the panel while preserving contrast.
- **Major grid** `#54d8e0`: unobtrusive but visible guides; minor grid lines are removed to reduce clutter.
- **Typography**: bold plot titles improve hierarchy and scannability.

When publishing, always verify contrast for your chosen geoms/colours against the background (e.g., dark text/lines on the light background typically meet WCAG AA).

# Reproducibility

```{r}
sessionInfo()
```
